---
title: "Getting started with gq"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with gq}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dpi = 150
)
```

**gq** is a style management system for cartography. Define your map styles once
in a JSON registry, then translate them to any rendering target ŌĆö tmap, mapgl,
leaflet, ggplot2. Change a color in one place and every map updates.

## The problem

If you make maps in R *and* QGIS *and* the web, you're maintaining the same
colors, line weights, and classification breaks in multiple places. Change a
stream color in QGIS? Now you need to update your tmap code, your leaflet code,
and your MapLibre style. It doesn't scale.

## The solution

A canonical style registry that serves as the single source of truth:

```
QGIS Project (.qgs)
  Ōåō gq_qgs_extract()
registry.json
  Ōåō gq_tmap_style() / gq_mapgl_style()
tmap, mapgl, leaflet, ggplot2
```

## Load a registry

The registry is just JSON. Load it with `gq_registry_read()`:

```{r load-registry}
library(gq)

reg_path <- system.file("examples", "demo_registry.json", package = "gq")
reg <- gq_registry_read(reg_path)

names(reg$layers)
```

Each layer has a type (polygon, line, point) and style properties:

```{r inspect-layer}
reg$layers$lake
```

## Translate to tmap

`gq_tmap_style()` converts a registry layer to tmap v4 arguments ŌĆö ready to
pass directly to `tm_polygons()`, `tm_lines()`, or `tm_dots()`:

```{r tmap-args}
gq_tmap_style(reg$layers$lake)

gq_tmap_style(reg$layers$stream)
```

## Build a map

gq ships with small demo datasets (`watershed`, `lake`, `stream`, `road`,
`crossing`) for illustration. Here's a full map styled entirely from the
registry:

```{r demo-map, fig.alt = "Map showing watersheds, lakes, streams, roads and crossings styled from the gq registry"}
library(tmap)
library(sf)

# Load demo spatial data
data(watershed, lake, stream, road, crossing, package = "gq")

# Style everything from the registry ŌĆö one source of truth
tm_shape(watershed) +
  do.call(tm_polygons, gq_tmap_style(reg$layers$watershed)) +
tm_shape(lake) +
  do.call(tm_polygons, gq_tmap_style(reg$layers$lake)) +
tm_shape(stream) +
  do.call(tm_lines, gq_tmap_style(reg$layers$stream)) +
tm_shape(road) +
  tm_lines(col = "grey50", lwd = 1.5) +
tm_shape(crossing) +
  do.call(tm_dots, gq_tmap_style(reg$layers$crossing))
```

Every color, opacity, and line weight comes from `reg` ŌĆö not hard-coded in the
plotting code.

## Classified layers

Many layers use categorized symbology (road types, fish passage status). The
registry stores per-class colors:

```{r classification}
reg$layers$road$classification
```

`gq_tmap_classes()` extracts this into a format ready for `tm_scale_categorical()`:

```{r tmap-classes}
road_cls <- gq_tmap_classes(reg$layers$road)
road_cls
```

```{r classified-map, fig.alt = "Map showing roads colored by type and crossings colored by fish passage status"}
crossing_cls <- gq_tmap_classes(reg$layers$crossing)

tm_shape(watershed) +
  do.call(tm_polygons, gq_tmap_style(reg$layers$watershed)) +
tm_shape(lake) +
  do.call(tm_polygons, gq_tmap_style(reg$layers$lake)) +
tm_shape(stream) +
  do.call(tm_lines, gq_tmap_style(reg$layers$stream)) +
tm_shape(road) +
  tm_lines(
    col = road_cls$field,
    col.scale = tm_scale_categorical(
      values = road_cls$values,
      labels = road_cls$labels
    ),
    lwd = 2
  ) +
tm_shape(crossing) +
  tm_dots(
    fill = crossing_cls$field,
    fill.scale = tm_scale_categorical(
      values = crossing_cls$values,
      labels = crossing_cls$labels
    ),
    size = 0.5
  )
```

## MapLibre GL output

The same registry translates to MapLibre GL paint properties for web maps:

```{r mapgl-output}
gq_mapgl_style(reg$layers$lake)

gq_mapgl_style(reg$layers$stream)
```

For classified layers, `gq_mapgl_classes()` builds a MapLibre match expression:

```{r mapgl-classes}
str(gq_mapgl_classes(reg$layers$road))
```

This expression goes directly into `mapgl::add_line_layer(paint = list("line-color" = expr))`.

## Extract from QGIS

If you have an existing QGIS project, `gq_qgs_extract()` parses the .qgs XML
and builds a registry from it ŌĆö no PyQGIS needed:

```{r extract}
qgs_path <- system.file("examples", "mini_project.qgs", package = "gq")
extracted <- gq_qgs_extract(qgs_path)

names(extracted$layers)
extracted$layers$lakes
```

Write it out as your registry:

```{r write-registry, eval = FALSE}
jsonlite::write_json(extracted, "registry.json", pretty = TRUE, auto_unbox = TRUE)
```

## One source of truth

The workflow is:

1. Design styles in **QGIS** (the best visual tool for cartography)
2. Extract with `gq_qgs_extract()` ŌåÆ `registry.json`
3. In R scripts: `gq_tmap_style()` for static maps, `gq_mapgl_style()` for web
4. Change a color in the registry ŌåÆ every map updates

No more copy-pasting hex codes across files.
