---
title: "Map composition with gq and tmap"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Map composition with gq and tmap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 9,
  dpi = 150
)
```

gq translates registry styles into tmap arguments. This vignette shows a
complete map composition — basemap, layers, legend, logo, keymap — following
New Graph cartographic conventions.

## Study area: Neexdzii Kwa subbasin

A subbasin of the Neexdzii Kwa (Upper Bulkley River) in the traditional
territory of the Wet'suwet'en, bounded by Johnny David Creek (downstream)
and Richfield Creek (upstream). ~212 km², pulled from the BC Freshwater Atlas
via [fresh](https://github.com/NewGraphEnvironment/fresh).

```{r load-data}
library(gq)
library(sf)
library(tmap)
library(maptiles)

sf_use_s2(FALSE)

data(neexdzii_wsd, neexdzii_streams, neexdzii_lakes, neexdzii_wetlands,
     neexdzii_roads, neexdzii_railway, neexdzii_bc, neexdzii_wsg,
     package = "gq")

cat("Watershed:", round(as.numeric(st_area(neexdzii_wsd)) / 10000), "ha\n")
cat("Streams:", nrow(neexdzii_streams), "segments\n")
cat("Lakes:", nrow(neexdzii_lakes), "\n")
cat("Wetlands:", nrow(neexdzii_wetlands), "\n")
cat("Roads:", nrow(neexdzii_roads), "\n")
```

## Load styles from the registry

One call to `gq_reg_main()` loads the master registry. Every color on the
map traces back to this registry — streams, roads, railway, lakes, wetlands.

```{r load-registry}
reg <- gq_reg_main()

# Extract colors and widths from classified layers for manual use
stream_col <- reg$layers$streams_all$classification$classes[["__empty__"]]$color
stream_lwd <- reg$layers$streams_all$classification$classes[["__empty__"]]$width

road_hwy <- reg$layers$roads_dra$classification$classes[["RF"]]
road_other <- reg$layers$roads_dra$classification$classes[["__empty__"]]

railway_col <- reg$layers$railway$stroke$color
railway_lwd <- reg$layers$railway$stroke$width
```

## Data prep

Filter streams by order for display, split roads by type for layered styling.

```{r data-prep}
# Streams: order >= 3 for display, >= 5 for labels
streams_display <- neexdzii_streams[neexdzii_streams$stream_order >= 3, ]

# Stream labels: dissolve named streams to single point per name
streams_named <- neexdzii_streams[
  !is.na(neexdzii_streams$gnis_name) & neexdzii_streams$stream_order >= 5, ]

if (nrow(streams_named) > 0) {
  stream_labels <- do.call(rbind, lapply(
    split(streams_named, streams_named$gnis_name),
    function(x) {
      combined <- st_union(x)
      pt <- st_point_on_surface(combined)
      st_sf(gnis_name = x$gnis_name[1], geometry = pt, crs = st_crs(x))
    }
  ))
} else {
  stream_labels <- streams_named[0, ]
}

# Roads: split highways from other roads for distinct styling
highways <- neexdzii_roads[neexdzii_roads$road_type %in%
  c("RF", "RH1", "RH2"), ]
roads_other <- neexdzii_roads[!neexdzii_roads$road_type %in%
  c("RF", "RH1", "RH2"), ]

# Lake labels
lakes_named <- neexdzii_lakes[!is.na(neexdzii_lakes$gnis_name_1) &
                                neexdzii_lakes$gnis_name_1 != "", ]
```

## Basemap: Positron x hillshade blend

Label-free raster basemap gives terrain relief without competing with our
own labels. At sub-watershed scale (zoom 10+), Positron-NoLabels blended
with hillshade works well.

```{r basemap}
bbox <- st_bbox(neexdzii_wsd)
y_pad <- (bbox["ymax"] - bbox["ymin"]) * 0.03
bbox["ymin"] <- bbox["ymin"] - y_pad
bbox["ymax"] <- bbox["ymax"] + y_pad
bbox_sf <- st_as_sfc(bbox) |> st_set_crs(4326)

positron <- get_tiles(bbox_sf, provider = "CartoDB.PositronNoLabels",
                      zoom = 10, crop = TRUE)
relief <- get_tiles(bbox_sf, provider = "Esri.WorldShadedRelief",
                    zoom = 10, crop = TRUE)
relief_rs <- terra::resample(relief, positron)
p_n <- positron / 255
r_g <- terra::mean(relief_rs) / 255
blended <- terra::clamp(p_n * (r_g ^ 0.5) * 255, lower = 0, upper = 255)
basemap_stars <- stars::st_as_stars(blended)
```

## Keymap inset

Small overview map showing the subbasin within the Bulkley/Morice watershed
groups and BC.

```{r keymap}
keymap <- tm_shape(neexdzii_bc) +
  tm_borders(col = "grey60", lwd = 0.5) +
tm_shape(neexdzii_wsg) +
  tm_polygons(fill = stream_col, fill_alpha = 0.5,
              col = "#1f78b4", lwd = 0.5) +
tm_shape(neexdzii_wsd) +
  tm_polygons(fill = "#ef4545", col = "#ef4545", lwd = 0.3) +
tm_layout(
  frame = TRUE,
  bg.color = "white",
  inner.margins = c(0.02, 0.02, 0.02, 0.02)
)
```

## Main map

Draw order matters: wetlands first, then streams, lakes on top of streams,
transport infrastructure, then labels last.

```{r map-composition, fig.alt = "Map of a Neexdzii Kwa subbasin showing streams, lakes, wetlands, roads, and railway styled from the gq registry"}
bb_box <- st_as_sfc(bbox, crs = st_crs(neexdzii_wsd))
logo_path <- system.file("logo", "nge_icon_200.png", package = "gq")

tmap_mode("plot")

# Polygon layers use do.call() with gq_tmap_style() directly
m <- tm_shape(basemap_stars) +
  tm_rgb() +
tm_shape(bb_box) +
  tm_borders(lwd = 0, col = NA) +
tm_shape(neexdzii_wsd) +
  do.call(tm_polygons, gq_tmap_style(reg$layers$watershed)) +
tm_shape(neexdzii_wetlands) +
  do.call(tm_polygons, gq_tmap_style(reg$layers$wetland)) +
tm_shape(streams_display) +
  tm_lines(col = stream_col, lwd = stream_lwd * 2) +
tm_shape(neexdzii_lakes) +
  do.call(tm_polygons, gq_tmap_style(reg$layers$lake))

# Lake labels
if (nrow(lakes_named) > 0) {
  m <- m + tm_shape(lakes_named) +
    tm_text("gnis_name_1", size = 0.5, col = "#1f78b4",
            fontface = "italic",
            options = opt_tm_text(shadow = TRUE))
}

# Transport — thinnest first, highways on top
# Colors from roads_dra classification in registry
if (nrow(roads_other) > 0) {
  m <- m + tm_shape(roads_other) +
    tm_lines(col = road_other$color, lwd = road_other$width * 2)
}
if (nrow(highways) > 0) {
  m <- m + tm_shape(highways) +
    tm_lines(col = road_hwy$color, lwd = road_hwy$width * 2)
}

# Railway overlay (black base + white dashed) — color from registry
if (nrow(neexdzii_railway) > 0) {
  m <- m + tm_shape(neexdzii_railway) +
    tm_lines(col = railway_col, lwd = railway_lwd * 2) +
  tm_shape(neexdzii_railway) +
    tm_lines(col = "white", lwd = railway_lwd, lty = "42")
}

# Stream labels
if (nrow(stream_labels) > 0) {
  m <- m + tm_shape(stream_labels) +
    tm_text("gnis_name", size = 0.45, fontface = "italic", col = "#1a5276",
            options = opt_tm_text(shadow = TRUE, remove_overlap = TRUE))
}

# Manual legends — colors pulled from registry, not hardcoded
m <- m +
tm_add_legend(
  type = "polygons",
  labels = c("Lake", "Wetland"),
  fill = c(
    reg$layers$lake$fill$color,
    reg$layers$wetland$fill$color
  )
) +
tm_add_legend(
  type = "lines",
  labels = c("Stream", "Highway", "Road", "Railway"),
  col = c(stream_col, road_hwy$color, road_other$color, railway_col),
  lwd = c(0.5, 1.3, 0.5, 0.8),
  lty = c("solid", "solid", "solid", "twodash")
)

# Layout: four-corner rule
# - Legend: bottom-left
# - Logo: top-right
# - Scalebar: bottom-center
# - Keymap: bottom-right (via grid viewport)
m <- m +
tm_scalebar(
  breaks = c(0, 2, 4, 6),
  text.size = 0.5,
  position = c("center", "bottom"),
  margins = c(0, 0, 0, 0)
) +
tm_logo(logo_path, position = c("right", "top"), height = 2.5,
        margins = c(0, 0, 0, 0)) +
tm_layout(
  frame = TRUE,
  frame.lwd = 0.5,
  asp = 0,
  legend.position = c("left", "bottom"),
  legend.frame = TRUE,
  legend.bg.color = "white",
  legend.bg.alpha = 0.85,
  legend.text.size = 0.55,
  legend.title.size = 0.65,
  inner.margins = c(0.04, 0.001, 0.001, 0.001),
  outer.margins = c(0.003, 0.003, 0.003, 0.003),
  meta.margins = 0
)

print(m)
print(keymap, vp = grid::viewport(
  x = 0.86, y = 0.12, width = 0.25, height = 0.22
))
```

Every color on this map traces back to `gq_reg_main()`:

- **Polygon layers** (watershed, lake, wetland) use `do.call(tm_polygons, gq_tmap_style(...))`
- **Line layers** (streams, roads, railway) pull colors from classified registry entries
- **Legend colors** reference the same registry objects — change a color once, every element updates
