---
title: "Map composition with gq and tmap"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Map composition with gq and tmap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 9,
  dpi = 150
)
```

gq translates registry styles into tmap arguments. This vignette shows a
complete map composition — basemap, layers, legend, logo, keymap — following
New Graph cartographic conventions.

## Study area: Neexdzii Kwa subbasin

A subbasin of the Neexdzii Kwa (Upper Bulkley River) in the traditional
territory of the Wet'suwet'en, bounded by Johnny David Creek (downstream)
and Richfield Creek (upstream). ~212 km², pulled from the BC Freshwater Atlas
via [fresh](https://github.com/NewGraphEnvironment/fresh).

```{r load-data}
library(gq)
library(sf)
library(tmap)
library(maptiles)

sf_use_s2(FALSE)

data(neexdzii_wsd, neexdzii_streams, neexdzii_habitat,
     neexdzii_lakes, neexdzii_wetlands,
     neexdzii_crossings, neexdzii_fish_obs, neexdzii_falls,
     neexdzii_roads, neexdzii_railway, neexdzii_bc, neexdzii_wsg,
     package = "gq")

cat("Watershed:", round(as.numeric(st_area(neexdzii_wsd)) / 10000), "ha\n")
cat("Streams:", nrow(neexdzii_streams), "| Habitat:", nrow(neexdzii_habitat), "\n")
cat("Crossings:", nrow(neexdzii_crossings), "| Fish obs:", nrow(neexdzii_fish_obs),
    "| Falls:", nrow(neexdzii_falls), "\n")
cat("Lakes:", nrow(neexdzii_lakes), "| Wetlands:", nrow(neexdzii_wetlands), "\n")
cat("Roads:", nrow(neexdzii_roads), "| Railway:", nrow(neexdzii_railway), "\n")
```

## Load styles from the registry

One call to `gq_reg_main()` loads the master registry. `gq_style()` resolves
any layer by name — no manual color extraction needed.

```{r load-registry}
reg <- gq_reg_main()

# gq_style() returns backend-agnostic style info by name
gq_style(reg, "lake")
gq_style(reg, "railway")

# gq_tmap_style() wraps gq_style() with tmap-specific args
# For classified layers it wires up tm_scale_categorical() automatically
gq_tmap_style(reg, "crossings_pscis_assessment")
```

## Data prep

Filter streams by order and build label points. Where the bundled data
comes from a different source than the registry (e.g., bcfishpass vs WHSE),
the `field` parameter in `gq_tmap_style()` maps the alternative column name
to the same style — no column renames needed. See `inst/registry/xref_layers.csv`
for the cross-reference.

```{r data-prep}
# Streams: order >= 3 for display, >= 5 for labels
streams_display <- neexdzii_streams[neexdzii_streams$stream_order >= 3, ]

# Stream labels: dissolve named streams to single point per name
streams_named <- neexdzii_streams[
  !is.na(neexdzii_streams$gnis_name) & neexdzii_streams$stream_order >= 5, ]

if (nrow(streams_named) > 0) {
  stream_labels <- do.call(rbind, lapply(
    split(streams_named, streams_named$gnis_name),
    function(x) {
      combined <- st_union(x)
      pt <- st_point_on_surface(combined)
      st_sf(gnis_name = x$gnis_name[1], geometry = pt, crs = st_crs(x))
    }
  ))
} else {
  stream_labels <- streams_named[0, ]
}

# Lake labels
lakes_named <- neexdzii_lakes[!is.na(neexdzii_lakes$gnis_name_1) &
                                neexdzii_lakes$gnis_name_1 != "", ]
```

## Basemap: Positron x hillshade blend

Label-free raster basemap gives terrain relief without competing with our
own labels. At sub-watershed scale (zoom 10+), Positron-NoLabels blended
with hillshade works well.

```{r basemap}
# Pad the bbox to the canvas aspect ratio (7:9) so the map fills the page. The
# latitude correction a geographic CRS needs is applied from the CRS rather
# than by hand -- this data is lat/lon, so it fires here and would not in BC
# Albers.
bbox <- gq_bbox_aspect(neexdzii_wsd, asp = 7 / 9)

positron <- gq_basemap_tiles(bbox, provider = "CartoDB.PositronNoLabels",
                             zoom = 10, pad = 0, crs = NULL)
relief <- gq_basemap_tiles(bbox, provider = "Esri.WorldShadedRelief",
                           zoom = 10, pad = 0, crs = NULL)

# gq_basemap_tiles() returns NULL on a failed fetch so the caller can draw an
# unshaded map rather than lose the figure. Honour that here rather than only
# documenting it -- a tile hiccup during a pkgdown build should not fail it.
basemap_stars <- if (is.null(positron) || is.null(relief)) {
  NULL
} else {
  # A relief tile service spans the full range, so the gamma operator suits it.
  # A hillshade derived from a DEM is far more contrasty and wants
  # `method = "weight"`, which caps how much brightness can be removed.
  gq_basemap_blend(positron, relief, method = "gamma", gamma = 0.5)
}

# Now that the frame is known, drop anything outside it. gq_bbox_clip() returns
# NULL rather than a zero-row object, which is what tm_shape() needs -- it
# errors on an empty geometry set instead of skipping it.
streams_display <- gq_bbox_clip(streams_display, bbox)
```

## Keymap inset

Small overview map showing the subbasin within the Bulkley/Morice watershed
groups and BC.

```{r keymap}
# Watershed group fill matches lake stroke color from registry
lake_sty <- gq_style(reg, "lake")

keymap <- tm_shape(neexdzii_bc) +
  tm_borders(col = "grey60", lwd = 0.5) +
tm_shape(neexdzii_wsg) +
  tm_polygons(fill = lake_sty$stroke$color, fill_alpha = 0.5,
              col = lake_sty$stroke$color, lwd = 0.5) +
tm_shape(neexdzii_wsd) +
  tm_polygons(fill = "#ef4545", col = "#ef4545", lwd = 0.3) +
tm_layout(
  frame = TRUE,
  bg.color = "white",
  inner.margins = c(0.02, 0.02, 0.02, 0.02)
)
```

## Main map

Draw order matters: polygons first, then habitat lines, base streams, lakes
on top, transport, point features (crossings, fish, falls), then labels last.

```{r map-composition, fig.alt = "Map of a Neexdzii Kwa subbasin showing salmon habitat, crossings, fish observations, streams, lakes, wetlands, roads, and railway styled from the gq registry"}
bb_box <- st_as_sfc(bbox, crs = st_crs(neexdzii_wsd))
logo_path <- system.file("logo", "nge_icon_200.png", package = "gq")

tmap_mode("plot")

# Pull styles from registry — all colors trace back to gq_reg_main()
stream_sty <- gq_style(reg, "streams_all")
railway_sty <- gq_style(reg, "railway")
fish_sty <- gq_style(reg, "bcfishobs_fiss_fish_observations")
falls_sty <- gq_style(reg, "fiss_obstacles")
lake_sty <- gq_style(reg, "lake")

# Polygon layers — do.call() with gq_tmap_style() directly
m <- tm_shape(basemap_stars) +
  tm_rgb() +
tm_shape(bb_box) +
  tm_borders(lwd = 0, col = NA) +
tm_shape(neexdzii_wsd) +
  tm_polygons(fill_alpha = 0, col = "#2c3e50", lwd = 1.5) +
tm_shape(neexdzii_wetlands) +
  do.call(tm_polygons, gq_tmap_style(reg, "wetland"))

# Salmon habitat — classified by mapping_code (xref: streams_salmon_vw uses
# mapping_code, registry expects mapping_code_salmon from WHSE source)
m <- m +
tm_shape(neexdzii_habitat) +
  do.call(tm_lines, gq_tmap_style(reg, "streams_salmon", field = "mapping_code"))

# Base streams on top of habitat — simple style, width scaled up for display.
# gq_bbox_clip() returns NULL when nothing survives the frame, so test for it
# rather than handing tm_shape() an empty geometry set — the same guard the
# basemap uses above, and the one the file already applies to railway and falls.
if (!is.null(streams_display)) {
  m <- m +
    tm_shape(streams_display) +
    tm_lines(col = stream_sty$classification$values[[1]],
             lwd = stream_sty$classification$widths[[1]] * 2)
}
m <- m +
tm_shape(neexdzii_lakes) +
  do.call(tm_polygons, gq_tmap_style(reg, "lake"))

# Lake labels — color from registry
if (nrow(lakes_named) > 0) {
  m <- m + tm_shape(lakes_named) +
    tm_text("gnis_name_1", size = 0.5, col = lake_sty$stroke$color,
            fontface = "italic",
            options = opt_tm_text(shadow = TRUE))
}

# Roads — classified by transport_line_type_code, bundled data column is
# road_type (xref: alias from data-raw query)
m <- m +
tm_shape(neexdzii_roads) +
  do.call(tm_lines, gq_tmap_style(reg, "roads_dra", field = "road_type"))

# Railway — base + white dashed overlay, colors from registry
if (nrow(neexdzii_railway) > 0) {
  m <- m + tm_shape(neexdzii_railway) +
    tm_lines(col = railway_sty$stroke$color,
             lwd = railway_sty$stroke$width * 2) +
  tm_shape(neexdzii_railway) +
    tm_lines(col = "white", lwd = railway_sty$stroke$width, lty = "42")
}

# Crossings — classified by barrier_status (xref: bcfishpass.crossings uses
# barrier_status, registry expects barrier_result_code from PSCIS WHSE)
m <- m +
tm_shape(neexdzii_crossings) +
  do.call(tm_dots, gq_tmap_style(reg, "crossings_pscis_assessment",
                                  field = "barrier_status"))

# Fish observations — shape and color from registry
if (nrow(neexdzii_fish_obs) > 0) {
  m <- m + tm_shape(neexdzii_fish_obs) +
    tm_symbols(shape = 24, fill = fish_sty$mark$color,
               col = fish_sty$mark$color, size = 0.12)
}

# Falls — shape and color from registry
if (nrow(neexdzii_falls) > 0) {
  m <- m + tm_shape(neexdzii_falls) +
    tm_symbols(shape = 22, fill = falls_sty$mark$color,
               col = falls_sty$mark$color, size = 0.2)
}

# Stream labels
if (nrow(stream_labels) > 0) {
  m <- m + tm_shape(stream_labels) +
    tm_text("gnis_name", size = 0.45, fontface = "italic", col = "#1a5276",
            options = opt_tm_text(shadow = TRUE, remove_overlap = TRUE))
}

# Legends from the registry. gq_tmap_legend() partitions by geometry type,
# expands classified layers to one entry per class, merges simple and classified
# layers of the same type into one legend, and collapses classes that render
# identically. `present` cuts road classes to the ones the data actually has.
leg <- gq_tmap_legend(
  reg,
  c("Lake" = "lake", "Wetland" = "wetland", "Stream" = "streams_all",
    "roads_dra", "Railway" = "railway",
    "crossings_pscis_assessment",
    "Fish obs" = "bcfishobs_fiss_fish_observations",
    "Falls" = "fiss_obstacles"),
  present = list(roads_dra = unique(neexdzii_roads$road_type))
)

m <- Reduce(`+`, c(list(m), lapply(leg, function(x) do.call(tm_add_legend, x))))

# Layout: four-corner rule
# - Legend: bottom-left
# - Logo: top-right
# - Scalebar: bottom-center
# - Keymap: bottom-right (via grid viewport)
m <- m +
tm_scalebar(
  # gq_scale_breaks() needs metre units and says so, so project the bbox first
  breaks = gq_scale_breaks(st_bbox(st_transform(st_as_sfc(bbox), 3005))),
  text.size = 0.5,
  position = c("center", "bottom"),
  margins = c(0, 0, 0, 0)
) +
tm_logo(logo_path, position = c("right", "top"), height = 2.5,
        margins = c(0, 0, 0, 0)) +
tm_layout(
  frame = TRUE,
  frame.lwd = 0.5,
  asp = 0,
  legend.position = c("left", "bottom"),
  legend.frame = TRUE,
  legend.bg.color = "white",
  legend.bg.alpha = 0.85,
  legend.text.size = 0.5,
  legend.title.size = 0.6,
  inner.margins = c(0.001, 0.001, 0.001, 0.001),
  outer.margins = c(0.003, 0.003, 0.003, 0.003),
  meta.margins = 0
)

print(m)
# gq_tmap_keymap() supplies the PLACEMENT here, not the map: it derives the
# viewport centre from the corner and margin and sizes it off the canvas aspect,
# which is the arithmetic every hand-rolled inset hardcodes. Its own two-layer
# map is discarded in favour of `keymap` above, which carries a third layer
# (watershed groups) that the (aoi, context) signature cannot express.
km <- gq_tmap_keymap(neexdzii_wsd, neexdzii_bc, asp = 7 / 9)
print(keymap, vp = km$viewport)
```

Every color on this map traces back to `gq_reg_main()`:

- **Simple layers** (lake, wetland, streams, railway) use `gq_tmap_style(reg, "name")` directly
- **Classified layers** (crossings, roads, habitat) use `do.call()` with `gq_tmap_style()` — classification field, colors, and labels all come from the registry
- **The legend** is built by `gq_tmap_legend()` from the registry — it partitions
  by geometry type, expands classified layers, collapses classes that render
  identically, and cuts road classes to the ones present in the data. Change a
  colour in the registry and every element updates
